\chapter{Quadratic Forms over $\Q$ and $\Q_p$}

\section{Quadratic Forms}\label{sec:quadratic_form}

The following definition is available in \mathlib under the name \code{QuadraticForm}:
\begin{definition}\label{def:quadratic_form}\leanok
Let $M$ be a module over a commutative ring $A$. A function $f : M \to A$ is called a \emph{quadratic form} on $M$ if 
\begin{enumerate}
\item $f(ax) = a^2f(x)$ for all $a \in A, x \in M$,
\item The function $(x, y) \mapsto f(x + y) - f(x) - f(y)$ is a bilinear form.
\end{enumerate}
\end{definition}

The \code{QuadraticForm} folder of \mathlib contains definitions and results about quadratic forms, including the bilinear map associated to a quadratic form (see \code{QuadraticMap.associated}) and the definition and characterization of nondegenerate quadratic form (see \code{QuadraticMap.Nondegenerate}).

\begin{definition}\label{def:isotropic}
\lean{QuadraticForm.Isotropic}
\uses{def:quadratic_form}
\leanok
A quadratic form $f : M \to A$ is called \emph{isotropic} if there exists a nonzero $x \in M$ such that $f(x) = 0$.
\end{definition}

\section{Quadratic Forms over $\Q_p$}\label{sec:quadratic_form_padic}

\begin{definition}\label{def:discr}
	\uses{def:Hilbert_symbol}
	Let $p$ be a prime number and let $f$ be a quadratic form of rank $n$ over $\Q_p$. Take a diagonal quadratic form $a_1X_1^2 + \cdots + a_nX_n^2$ which is equivalent to $f$; then the \emph{discriminant} of $f$ is the element $d(f) = a_1 \cdots a_n \in \Q_p^\times/{\Q_p^\times}^2$.
	
Note that \mathlib contains \code{QuadraticMap.discr}, as an element of $\Q_p$.
\end{definition}

\begin{definition}\label{def:HM_inv}
\uses{def:Hilbert_symbol}
Let $p$ be a prime number and let $f$ be a quadratic form of rank $n$ over $\Q_p$. Take a diagonal quadratic form $a_1X_1^2 + \cdots + a_nX_n^2$ which is equivalent to $f$; then the \emph{Hasse--Minkowski invariant} of $f$ is the element
\[ \epsilon(f) := \prod_{i < j}(a_i, a_j).\]
\end{definition}

\begin{lemma}\label{lem:HM_inv_well_defined}
\uses{def:HM_inv}
The value $\epsilon(f)$ does not depend on the choice of diagonal quadratic form equivalent to $f$.
\end{lemma}
\begin{proof}
	
\end{proof}

\begin{lemma}\label{lem:HM_inv_pm_one}
\uses{def:HM_inv}
For every quadratic form $f$, we have $\epsilon(f) = \pm 1$.
\end{lemma}
\begin{proof}
Each term in the product is equal to $\pm 1$.
\end{proof}


\section{Hasse--Minkowski}\label{sec:Hasse--Minkowski}

The goal of this section is to prove the Hasse--Minkowski theorem over the rational numbers. To simplify the formalization process, rather than including a single long proof of this theorem, we will split it into smaller lemmas that can be formalized independently.

Denote by $V$ the union of the set of prime numbers and the symbol $\infty$, and denote $\Q_\infty = \R$. 

Let $f$ be a quadratic form over $\Q$. For each $v \in V$, the injection $\Q \to \Q_v$ allows one to view $f$ as a quadratic form over $\Q_v$, which we will denote $f_v$.

\begin{definition}\label{def:everywhere_locally_isotropic}
\lean{QuadraticForm.EverywhereLocallyIsotropic}
\uses{def:isotropic}
\leanok
A quadratic form $f$ with coefficients in $\Q$ is \emph{everywhere locally isotropic} if $f_v$ is isotropic for all $v \in V$.
\end{definition}

\begin{theorem}[Hasse--Minkowski]\label{thm:Hasse--Minkowski}
\lean{QuadraticForm.HasseMinkowski}
\uses{lem:HM_of_degenerate, lem:ev_loc_isotr_of_isotr}
\leanok
Let $f$ be a quadratic form over $\Q$. Then $f$ is isotropic if and only if $f$ is everywhere locally isotropic. In other words, $f$ has a global zero if and only if $f$ has everywhere a local zero. 
\end{theorem}
\begin{proof}
Follows from the results below.
\end{proof}

\begin{theorem}[Hasse--Minkowski]\label{lem:HM_of_degenerate}
	\lean{QuadraticForm.HasseMinkowski}
	\uses{lem:ev_loc_isotr_of_isotr, lem:isotr_of_rk_zero, lem:isotr_of_rk_one, lem:isotr_of_rk_two, lem:isotr_of_rk_three, lem:isotr_of_rk_four, lem:isotr_of_fiv_le_rk}
	\leanok
	Let $f$ be a degenerate quadratic form over $\Q$. Then $f$ is isotropic if and only if $f$ is everywhere locally isotropic. In other words, $f$ has a global zero if and only if $f$ has everywhere a local zero. 
\end{theorem}
\begin{proof}
	
\end{proof}

From now on, given \ref{lem:HM_of_degenerate}, we may assume that $f$ is a nondegenerate quadratic form.

\begin{lemma}\label{lem:ev_loc_isotr_of_isotr}
\lean{QuadraticForm.Isotropic.everywhereLocallyIsotropic}
\uses{def:everywhere_locally_isotropic}
\leanok
Let $f$ be an isotropic quadratic form over $\Q$. Then $f$ is everywhere locally isotropic.
\end{lemma}
\begin{proof}
	
\end{proof}

\begin{lemma}\label{lem:isotr_of_rk_zero}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_rank_zero}
	\uses{def:everywhere_locally_isotropic}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank zero which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
Suffices to see that every rank zero quadratic form over a field is anisotropic.
\end{proof}

\begin{lemma}\label{lem:equiv_sum_squares}
\lean{QuadraticForm.equivalent_weightedSumSquares_units_of_nondegenerate}
\uses{def:quadratic_form}
\leanok
Let $K$ be a field of characteristic not equal to $2$, let $M$ be a $K$-vector space of positive dimension $n$ and let $f : M \to K$ be a
nondegenerate quadratic form. Then $f$ is equivalent to a form $X_1^2 +  a_2X_2^2 + \cdots + a_nX_n^2$, for some $a_i \in \Q^\times$.
\end{lemma}
\begin{proof}
Note that \mathlib contains a similar statement\footnote{\code{QuadraticForm.equivalent_weightedSumSquares_units_of_nondegenerate}}, where the coefficient of $X_1$ is not assumed to be $1$.
\end{proof}

\begin{lemma}\label{lem:isotr_of_rk_one}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_rank_one}
	\uses{def:everywhere_locally_isotropic}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank one which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
Suffices to see that every rank zero quadratic form of rank one over a field is anisotropic.
\end{proof}

\begin{lemma}\label{lem:isotr_of_rk_two}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_rank_two}
	\uses{def:everywhere_locally_isotropic, lem:equiv_sum_squares}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank two which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
$f$ is equivalent to $X_1^2 - aX_2^2$, for some nonzero $a$, which is positive because $f_\infty$ represents $0$. Write
\[ a = \prod_p p^{v_p(a)}.\]
Since $f_p$ represents $0$, $a$ is a square in $\Q_p$, so $v_p(a)$ is even. It follows that $a$ is a square in $\Q$ and $f$ represents $0$.
\end{proof}

\begin{lemma}\label{lem:isotr_of_rk_three}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_rank_three}
	\uses{def:everywhere_locally_isotropic, lem:equiv_sum_squares}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank three which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
	
\end{proof}

\begin{lemma}\label{lem:isotr_of_rk_four}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_rank_four}
	\uses{def:everywhere_locally_isotropic, lem:equiv_sum_squares}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank four which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
	
\end{proof}

\begin{lemma}\label{lem:isotr_of_five_le_rk}
	\lean{QuadraticForm.EverywhereLocallyIsotropic.isotropic_of_five_le_rank}
	\uses{def:everywhere_locally_isotropic, lem:equiv_sum_squares, thm:padic_squares_open, lem:approx_thm}
	\leanok
	Let $f$ be a quadratic form over $\Q$ of rank $n \ge 5$ which is everywhere locally isotropic. Then $f$ is isotropic.
\end{lemma}
\begin{proof}
The proof uses induction on $n$. Write $f = h - g$, where $h = a_1X_1^2 + a_2X_2^2$, $g = -(a_3X_3^2 + \cdots + a_nX_n^2)$.
Let $S$ be the finite subset of $V$ consisting of $\infty, 2$, and the primes $p$ such that $v_p(a_i) \ne 0$ for $i \ge 3$.
Let $v \in S$. Since $f_v$ represents $0$, there exists $c_v \in \Q_v^\times$ which is represented in $\Q_v$ by $h$ and $g$, that is, there exist $x_i^v \in \Q_v$, $i=1, \dots, n$ (with $x_i^v \ne 0$ for $i=1, 2$, and some $i \ge 3$) such that 
\[ h(x_1^v, x_2^v) = c_v = g(x_3^v, \cdots, x_n^v).\]
By \ref{thm:padic_squares_open}, the squares of $\Q_p^\times$ form an open set. Combining this with the approximation theorem \ref{lem:approx_thm}, we deduce the existence of $x_1, x_2 \in \Q^\times$ such that, denoting $c := h(x_1, x_2)$, one has $c/c_v \in {\Q_v^\times}^2$ for all $v \in S$.

Consider the form $f_1 := cZ^2 - g$. For each $v \in S$, $g$ represents $c_v$ in $\Q_v$, and hence also $c$ because $c/c_v \in {\Q_v^\times}^2$, hence $f_1$ represents $0$ in $\Q_v$. For $v \notin S$, the coefficients $-a_3, \dots, -a_n$ of $g$ are $v$-adic units, so the discriminant $d_v(g)$ is a $v$-adic unit and $\epsilon_v(g)=1$.


In all cases, $f_q$ represents $0$ in all $\Q_v$, and since the rank of $f$ is $n-1$, the inductive hypothesis shows that $f_1$ represents $0$ in $\Q$, or equivalently, $g$ represents $c$ in $\Q$. But $h$ also represents $c$, so $f$ represents $0$.

\end{proof}